### 2009-11-24 docs/marcello-dissertation/thesis/chapter5-taxonomic-persistence-storage-selection-for-netbeams.tex
Updates to chapter 2 through 7.
Chapter 2 lost the KVP section, since this approach had not been identified in
the literature for sensor networks. In this way, this makes the primary finding
and contribution of the dissertation. It will only appear during the technology
selection, what is the direct representation of what had happened;

chapter 3 adds details to the taxonomies;

Chapter 4 was cut short to describe SF-BEAMS, NetBEAMS in separate, as well as
keeps the entire implementation to the last chapter (appendix). 

Chapter 5 is the selection of the technology, but still misses highlights

This chapter analyzes each of the taxonomies defined in chapter
\ref{chap:taxonomies} against the requirements defined by our case study,
NetBEAMS, in chapter \ref{sec:problem-requirements}. Moreover, it evaluates
different database technologies against on those taxonomies, and tries to select
one of them which can be a candidate for an experimental analysis.

First, the contender technologies are as follows:

\begin{itemize}
  \item MySQL
  \item Microsoft SQL 
  \item Berkeley TinyDB
  \item MongoDB
  \item CouchDB
  \item IBM DB2
\end{itemize}

\section{Analysis of the Purpose of Sensor Data Taxonomy}

Considering the infrastructure of NetBEAMS, as described in the previous
chapter, and the characteristics of the sensor network defined at the SF-BEAMS
\cite{sfbeams2006} site, the data characteristics are as follows:

@@

Given the facts collected from the observations of the execution of NetBEAMS
through the SF-BEAMS sensor network, we can say that the purpose of collected
data is \textbf{Data Archival}.

@@

Any of the contenders are options to cover for Data Archival.

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Microsoft SQL}: Supports
  \item \textbf{Berkeley TinyDB}: Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{IBM DB2}: Supports
\end{itemize}

@@

Based in the observations described, the type used for storing
data can be either the \textbf{External Storage} or the
\textbf{Data-Centric Storage}. The former is the current approach used by the
SF-BEAMS sensor network, while the latter adopted as an improvement of the
persistence infrastructure for NetBEAMS.

@@

In order to support this strategy, a technique called Database Sharding
\cite{db-shard02} is a technique used in distributed database systems that
uses denormalized data models to improve scalability in terms of data storage
and data location. It is a type of database system patitioning where
de-normalized data is spread across different other hosts, or shard \cite{db-shard01}.
Different models of database sharding can be developed \cite{db-shard-schemas}.

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Microsoft SQL}: Supports
  \item \textbf{Berkeley TinyDB}: Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{IBM DB2}: Supports
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Microsoft SQL}: Supports
  \item \textbf{Berkeley TinyDB}: Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{IBM DB2}: Supports
\end{itemize}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.7]{../diagrams/KVP-on-Relational-Model}
  \caption{KVP Data Model implementation using Relational Model}
  \label{fig:KVP-on-Relational-Model}
\end{figure}

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.7]{../diagrams/persistence-example-relational-kvp}
  \caption{KVP over Relational Model Database Prototype}
  \label{fig:persistence-example-relational-kvp}
\end{figure}

@@

The advance of Cloud Computing, and dynamic web applications have motivated the
database community to implement different mechanisms for persisting data
without the heavy-process of data modeling. Given the fact that sensor networks
may be updated at any time by the addition of new sensor types, the use of
moding approaches such as Key-Value Database have been proposed \cite{db-kvp}.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/persistence-example-kvp}
  \caption{KVP Database instance Prototype}
  \label{fig:persistence-example-kvp}
\end{figure}

@@

The data models were compared by \cite{db-is-rdbs-dommed} and can be summarized
as follows:

@@

\begin{table}
    \label{tab:ysi-data-distribution}
    \caption{Schema-Dependent X Schema-less Databases: Data Access}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item The basic operations CRUD \footnote{Create-Retrieve-Update-Update
      database operations} data are performed using a structured language such
      as the SQL or XPath;
      \item Query languages can access data from different tables through
      joins, contains functions for aggregation and complex filter.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item The CRUD operations are performed via API \footnote{Application
      Programming Interface} through programming languages;
      \item Some technologies provide basic SQL-like syntax for filter criteria
      with some predicates like =, !=, <, > that ca be applied;
      \item The data and application integrity logic is placed in the
      application layer.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

@@

\begin{itemize}
  \item MySQL: NO SUPPORT
  \item Microsoft SQL: NO SUPPORT
  \item Berkeley TinyDB: NO SUPPORT
  \item \textbf{MongoDB}: Supports Both
  \item \textbf{CouchDB}: Supports Both
  \item IBM DB2: NO SUPPORT
\end{itemize}

@@

Sensor Network data can be saved in either Centralized or Distributed Systems.
As described in  While Centralized Database Systems are easier to manage, it may
face challenges regarding its data. For this reason, the use of partitioning or
sharding would be an addition to persist sensor data.

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Microsoft SQL}: Supports
  \item \textbf{Berkeley TinyDB}: NO Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{IBM DB2}: Supports
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Microsoft SQL}: Supports
  \item \textbf{Berkeley TinyDB}: Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{IBM DB2}: Supports
\end{itemize}

@@

\begin{table}
    \label{tab:ysi-data-distribution}
    \caption{Amount of data produced by the RTC's YSI sondes}
        \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|}\hline 
        \textbf{Database} & \textbf{MySQL} & \textbf{MicrosoftSQL} & \textbf{TinyDB} &
        \textbf{MongoDB} & \textbf{CouchDB} & \textbf{IBM DB2}\\\hline 
        Centralized Query & + & + & + & + & + & + \\\hline 
        Distributed System & ++ & ++ & - & ++ & ++ & +\\\hline 
        Schema-less & - & - & - & + & + & +\\\hline 
        Provenance Support & + & + & + & + & + & +\\\hline 
        NO-SQL Query & - & - & - & + & + & +\\\hline 
        Data Partition & + & + & - & + & + & +\\\hline 
        Hot Backup & + & + & - & - & - & +\\\hline 
        Export Capability & + & + & - & + & + & -\\\hline 
        Programming Lang. & + & + & + & + & + & +\\\hline
        Script Lang. & + & + & - & + & + & +\\\hline
        Open-Source & + & - & + & + & + & -\\\hline
        \end{tabular}
        \end{center}
\end{table}

>>>>

This chapter analyzes each of the taxonomies defined in chapter
\ref{chap:taxonomies} against the requirements defined by NetBEAMS as our case
study in chapter \ref{sec:problem-requirements}, following the scope of a given
solution also defined in that chapter. The contenders were drawn from a mixed
list of technologies either related in the literature review of chapter 2 or
from the Internet.

\section{Database Systems Contenders}

This section describes the database systems contenders considered to be used
for the implementation of Data Persistence layer for NetBEAMS. As described in
the previous chapter, the selection of the technologies must be based on the
list of functional and non-functional requirements specified within the scope
of implementation defined.

One of the primary questions related to data persistence is regarding the data
model and type of the database, such as the use of relational model or
structured models. In addition to that, and given the current time of academic
research and new trends distributed computing such as Cloud Computing and the
Internet \cite{cloud-comp-survey}, the selection of a data persistence for
sensor networks should maintain a wider spectrum of the types of database
systems. Among other intriguing questions, \cite{db-is-rdbs-dommed} reflects 
about the use of alternatives to the relational model for dynamic
environments such as data persistence for the Web. Similarly, these
trends are also related to the advances of Cloud Computing and its persistence
layer as a solution to dynamic systems \cite{cloud-comp-architectures}. For
this reason, this work also included the overview of alternative data models
for the development of an experiment that brings data persistence to NetBEAMS.

\begin{itemize}
  \item \textbf{MySQL}: an open-source relational database system that is the
  primary choice of the development of projects of different sizes. It was selected
  because of its popularity in the current community;
  \item \textbf{Oracle}: is the most popular commercial database system, having
  a campus wide academic licenses availabe;
  \item \textbf{TinyDB}\cite{tinydb}: as referred in the literature review,
  TinyDB is a database system used in many different sensor networks implementation
  \cite{sn-db-tinydb};
  \item \textbf{MongoDB}\cite{mongodb}: an open-source key-value database system
  described in the article that questions the use of the relational model
  \cite{db-is-rdbs-dommed}, as well as the Cloud Computing trends survey
  \cite{cloud-comp-architectures}.
  \item \textbf{DB2}\cite{db2}: in the search of an XML database system, DB2 is
  listed as a hybrid database system that offers both XML and relational model
  \cite{db-xml-enabled}. It was selected because NetBEAMS can represent the
  collected data in the XML format \ref{sec:dsp-message}.
\end{itemize}

The following sections analyzes the taxonomies defined in chapter 3, along with
the requirements for the Data Persistence Layer for NetBEAMS, in section 4.

\section{Analysis of the Purpose of Sensor Data}

Considering the infrastructure of NetBEAMS, as described in the previous
chapter, and the characteristics of the sensor network defined at the SF-BEAMS
site \cite{sfbeams2006}, the data characteristics are as follows:

@@

As summarized in the previous section, the nature of data from NetBEAMS through
SF-BEAMS is used for the purpose of \textbf{Data Archival}.

@@

As it is supposed to be, all the contenders are developed to support data
persistence and management, what is the requirement for the Data Archival
purpose of the sensor network data.

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Oracle}: Supports
  \item \textbf{TinyDB}: Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{DB2}: Supports
\end{itemize}

@@

Based in the observations also described in the previous chapter, the type used
for storing data can be either the \textbf{External Storage}. However, the use
of a \textbf{Data-Centric Storage} approach can be considered. The former is the
current approach used by the SF-BEAMS sensor network, while the latter can be
explored in the experiments.

@@

In order to support this strategy, approaches such as data partitioning based
on the entities called Database Sharding
\cite{db-shard02}, or the partitioning based on tables
\cite{db-table-partition}. The former is a technique used in distributed
database systems that uses denormalized data models to improve scalability in
terms of data storage and data location, while the latter defines partitioning
techniques in relational databases. It is a type of database system patitioning
where de-normalized data is spread across different other hosts, or shard
\cite{db-shard01}. Different models of database sharding can be developed
\cite{db-shard-schemas}.

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Oracle}: Supports
  \item \textbf{Berkeley TinyDB}: Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{DB2}: Supports
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Oracle}: Supports
  \item \textbf{Berkeley TinyDB}: Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{IBM DB2}: Supports
\end{itemize}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.6]{../diagrams/KVP-on-Relational-Model}
  \caption{KVP Data Model implementation using Relational Model}
  \label{fig:KVP-on-Relational-Model}
\end{figure}

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.75]{../diagrams/persistence-example-relational-kvp}
  \caption{KVP over Relational Model Database Prototype}
  \label{fig:persistence-example-relational-kvp}
\end{figure}

@@

The growth of distributed systems and the Internet have enabled the development
of powerful database servers that can be organized in the context of
infrastructure and how to represent models. An powerful up-coming approach is
the use of database systems that does not use structured language for querying
its data. One such example is called Key-Value Pair (KVP) data model
\cite{db-kvp}, which is also known as Document-Oriented, Distributed Hash
Table. Such a model has the following properties:

\begin{itemize}
  \item data is structured in collections of key-value pairs, as it is done in
  hash data structure. The definition of the key is a given property with its
  associating value;
  \item the query process is using mechanics similar to programming or
  scripting language, that is, the use of APIs in a given programming language.
\end{itemize}

There are no records of the use of this data model with Sensor Networks.
Different variants of such data model is the document-oriented data model,
where data is modeled using structured documents. However, these documents
have a dynamic structure that can be freely described without the use of a
schema validation mechanism as used in XML Schema. One example of such
approach is the use of the JSON data format \cite{json} in the implementation
of database systems which are becoming popular with the new trend called Cloud
Computing \cite{cloud-comp-architectures}. Databases implementing this
strategy is the mongoDB \cite{mongodb}.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.75]{../diagrams/persistence-example-kvp}
  \caption{KVP Database instance Prototype}
  \label{fig:persistence-example-kvp}
\end{figure}

@@

The relational and key-value data models were compared by
\cite{db-is-rdbs-dommed} and can be summarized as follows:

@@

\begin{table}
    \label{tab:ysi-data-distribution}
    \caption{Schema-Dependent X Schema-less Databases: Data Access}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item The basic operations CRUD\footnote{Create-Retrieve-Update-Update
      database operations} data are performed using a structured language such
      as the SQL or XPath;
      \item Query languages can access data from different tables through
      joins, contains functions for aggregation and complex filter.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item The CRUD operations are performed via API\footnote{Application
      Programming Interface} through programming languages;
      \item Some technologies provide basic SQL-like syntax for filter criteria
      with some predicates like =, !=, <, > that ca be applied;
      \item The data and application integrity logic is placed in the
      application layer.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

@@

\begin{itemize}
  \item MySQL: NO SUPPORT
  \item Oracle: NO SUPPORT
  \item Berkeley TinyDB: NO SUPPORT
  \item \textbf{MongoDB}: Supports Both
  \item IBM DB2: NO SUPPORT
\end{itemize}

@@

Sensor Network data can be saved in either Centralized or Distributed Systems.
While Centralized Database Systems are easier to manage, it may
face challenges regarding its data. In order to implement a Data-Centric
solution, a distribute database system must be in place in order to manage
the different data by a single node.

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Oracle}: Supports
  \item \textbf{Berkeley TinyDB}: NO Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{IBM DB2}: Supports
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Oracle}: Supports, Not Open-Source
  \item \textbf{Berkeley TinyDB}: Supports, Open-Source
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{IBM DB2}: Supports, Open-Source
\end{itemize}

@@

Each of the databases were scored according to its support to the different
types of taxonomies, and randomly selected by using the literature review's
list, as well as the Internet.

\begin{table}
    \label{tab:ysi-data-distribution}
    \caption{Amount of data produced by the RTC's YSI sondes}
        \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|}\hline 
        \textbf{Database} & \textbf{MySQL} & \textbf{Oracle} &
        \textbf{TinyDB} & \textbf{MongoDB} & \textbf{IBM DB2}\\\hline 
        Centralized Query & + & + & + & + & + \\\hline 
        Distributed System & ++ & ++ & - & ++ & +\\\hline 
        Schema-less & - & - & - & + & +\\\hline 
        Provenance Support & + & + & + & + & +\\\hline 
        NO-SQL Query & - & - & - & + & +\\\hline 
        Data Partition & + & + & - & + & +\\\hline 
        Hot Backup & + & + & - & - & +\\\hline 
        Export Capability & + & + & - & + & -\\\hline 
        Programming Lang. & + & + & + & + & +\\\hline
        Script Lang. & + & + & - & + & +\\\hline
        Open-Source & + & - & + & + & -\\\hline
        \end{tabular}
        \end{center}
\end{table}

### 2009-11-24 docs/marcello-dissertation/thesis/chapter6-data-persistence-component-design-and-architecture.tex
Updates to chapter 2 through 7.
Chapter 2 lost the KVP section, since this approach had not been identified in
the literature for sensor networks. In this way, this makes the primary finding
and contribution of the dissertation. It will only appear during the technology
selection, what is the direct representation of what had happened;

chapter 3 adds details to the taxonomies;

Chapter 4 was cut short to describe SF-BEAMS, NetBEAMS in separate, as well as
keeps the entire implementation to the last chapter (appendix). 

Chapter 5 is the selection of the technology, but still misses highlights

\section{Business Process Analysis}
\label{sec:business-process-analysis}
As described in section \ref{chap:netbeams-overview}, the DSP Platform was
designed to support an environmental sensor network based in San Francisco
Bay, composed by different types of sensors. The process of extracting the
measurements from the sensors to be reused using the DSP Platform can be
summarized following the UML Business Process Diagram \cite{uml} depicted in
figure \ref{fig:fig:dsp-persistence-business-process}. It is important to note
that this diagrams details remote communication between sensor devices and a
centralized server, as part of the main scenario to solve the problem of data
persistence.

@@

As a matter of fact, the data measurements collected from the sensor devices
are simple properties with keys and associated values as described in section
\ref{sec:sfbeams}, after data has been described. According to the descriptions
of the DSP Platform and the DSP Messages structure, the data is represented in
programming language formats such as Java \cite{java} or in a serialized
version in XML \cite{xml}. Similarly, the collected data contains measurements
that uniquely identifies them, together with other properties such as time
stamps of data creation and collected, as well as the the IP address
\footnote{Internet Protocol Address} of the DSP Client that produced the data.
Therefore, these are important properties for the proposed solution of this work.

@@

@@

As depicted by the the UML Sequence Diagram \cite{uml} of figure
\ref{fig:From-DSPWireTransport-Server-To-DSP-Broker}, a DSP Message
is delivery to the DSP Broker for its delivery to Data Consumers configured to
receive that specific message. For this reason, figre
\ref{fig:From-DSP-Broker-To-DSPDataPersistence-General-Sequence} continues the
steps of the DSP Broker from the gate "From DSP WireTransport Server to Broker
Gate", showing the steps taken to transfer a DSP Message to the Singleton
class TransientPersistenceLayer, responsible for maintaining a queue of
messages waiting to be sent to the database system.

@@

Next section details the implementation and deployment of the DSP Data
Persistence component, as well as the database technology chosen in the previous
section to provide the persistence for the collected data by the DSP Data
Persistence component.

>>>>

\section{Business Process Analysis}
\label{sec:business-process-analysis}
As described in section \ref{chap:netbeams-overview}, the DSP Platform was
designed to support an environmental sensor network based in San Francisco
Bay, composed by different types of sensors. The process of extracting the
measurements from the sensors to be reused using the DSP Platform can be
summarized following the UML Business Process Diagram \cite{uml} depicted in
figure \ref{fig:dsp-persistence-business-process}. It is important to note
that this diagrams details remote communication between sensor devices and a
centralized server, as part of the main scenario to solve the problem of data
persistence.

@@

As a matter of fact, the data measurements collected from the sensor devices
are simple properties with keys and associated values as described in section
\ref{sec:sfbeams}, after data has been described. According to the descriptions
of the DSP Platform and the DSP Messages structure, the data is represented in
programming language formats such as Java \cite{java} or in a serialized
version in XML \cite{xml}. Similarly, the collected data contains measurements
that uniquely identifies them, together with other properties such as time
stamps of data creation and collected, as well as the the IP
address\footnote{Internet Protocol Address} of the DSP Client that produced
the data. Therefore, these are important properties for the proposed solution
of this work.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.5]{../diagrams/DSP-Data-Persistence-Mongo-Classes}
  \caption{UML Class diagram with Persistence Service for MongoDB}
  \label{fig:DSP-Data-Persistence-Mongo-Classes}
\end{figure}

@@

As depicted by the the UML Sequence Diagram \cite{uml} of figure
\ref{fig:From-DSPWireTransport-Server-To-DSP-Broker}, a DSP Message
is delivery to the DSP Broker for its delivery to Data Consumers configured to
receive that specific message. For this reason, Figure
\ref{fig:From-DSP-Broker-To-DSPDataPersistence-General-Sequence} continues the
steps of the DSP Broker from the gate "From DSP WireTransport Server to Broker
Gate", showing the steps taken to transfer a DSP Message to the Singleton
class TransientPersistenceLayer, responsible for maintaining a queue of
messages waiting to be sent to the database system.

@@

\section{DSP Data Persistence Deployment}

The deployment of the persistence layer is comprised of two main
components: the DSP Data Persistence and the MongoDB Database Server. Two
different approaches can be taken for the deployment: a single server of a
cluster of mongoDB instances for a sharded environment.

The single server approach maintains all data located in a simple space.
Distributed Systems applications such as Master-Slave, Data Replication, etc
can be used in this approach for scalability
\ref{fig:DSP-Data-Persistence-Deployment-Single}.

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.7]{../diagrams/DSP-Data-Persistence-Deployment-Single}
  \caption{UML Deployment diagram with a single server}
  \label{fig:DSP-Data-Persistence-Deployment-Single}
\end{figure}

However, given that mongoDB provides support to Shards, then the implementation
of a Data-Centric approach can be done by deploying different mongoDB instances
in a distributed cluster, where each node can hold a given type of data. The
specification of what types of data should be stored in each of the cluster
nodes is defined by a shard-key.

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.7]{../diagrams/DSP-Data-Persistence-Deployment-Sharded}
  \caption{UML Deployment Diagram for a sharded server}
  \label{fig:DSP-Data-Persistence-Deployment-Sharded}
\end{figure}

More details about the implementation of the DSP Data Persistence can be see in
the next chapter.
